---
title: "QMEE_permutation"
author: "Ali"
date: "23/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## class 9
Date: Feb 23, 2021

- Permutation approach; no assumption, the weakness is to get confidence interval
  - the simplest way to hypo test is, if the 2 groups are the same we sample from them equally!
    - **counting the ant colonies;** 210 permutation
  
  - arithmatic mean to geometric is equivalent to go from linear to log transfor; think multiplicatively!
  - **Pond nutrients;** 
  - Monte Carlo; 
  
  ```{r 23Feb2021}

library("ggplot2"); theme_set(theme_bw())
library("lmPerm")
library("coin")
library("gtools")
  
```

## class 10
Date: Feb 25, 2021

permutation approach: A basically doing null-hyp testing in frequentists setup
you can do usually t-test or ANOVA instead of permutation test. But the permutation test is more general with less assumptions, more flexible, as long as you can think about scrambling the data based on treating the groups the same.




